---
title: Visualisation with ggplot
author: Johan Blixt Dackhammar
date: 2025-11-26
output: pdf_document
knitr: 
    opts_chunk: 
      echo: false
      warning: false
      message: false
---

```{r setup, echo=FALSE, include=FALSE}
library(tidyverse)
```

# Boxplots with violins

We start by loading and looking at the data.

```{r plot1, echo=TRUE}
data1 <- read_csv("../data/data_plot1.csv")
glimpse(data1)
```

We can see that the first unnamed column is row number and the last column "myaxis" is the combination of `name`, newline `\n` and n=`n`. Looking at the original graph, this equals to the axis labels. The graph uses overlapped violin and boxplots with a fill that seems to use the default color scheme. There is a title and y-axis label but no x-axis label or legend.

```{r}
plot1 <- data1 |> 
    ggplot(aes(myaxis, value, fill = name)) +
    geom_violin(show.legend = F, ) +
    geom_boxplot(show.legend = F, outliers = F, width = 0.1) +
    labs(
        title = "Boxplot + Violin", 
        y = "Value",
        x = ""
    ) +
    theme_bw() +
    theme(panel.grid.minor = element_blank())

ggsave("../output/plot1.pdf", plot = plot1, width = 7, height = 4, dpi = 300)
```

# Simple Scatter Plot

Here we are to recreate a scatterplot with area on x-axis and population on y-axis. Population density is mapped to point size and state to point color. There is a regression line. There is an added title, subtitle and axis labels for x and y axes but legends are not renamed. Theme looks like theme_bw()

```{r, echo=FALSE}
data2 <- read_csv("../data/data_plot2.csv") 
glimpse(data2)
plot2 <- data2 |> 
    ggplot(aes(area,poptotal)) +
    geom_point(aes(color = state, size = popdensity), position = "jitter") +
    geom_smooth(se = F) +
    labs(
        title = "Scatterplot",
        subtitle = "Area vs Population",
        y = "Population",
        x = "Area"
    ) +
    scale_x_continuous(limits = c(0,0.1), breaks = seq(0,0.1,0.025)) +
    scale_y_continuous(limits = c(0,500000), breaks = seq(0,500000,100000), labels = \(x) format(x, big.mark = "", scientific = F)) +
    theme_bw()
plot2
ggsave("../output/plot2.pdf", plot = plot2, width = 7, height = 4, dpi = 300)

```

# Counts plot

```{r}
data3 <- read_csv("../data/data_plot3.csv")
glimpse(data3)
data3 |> 
    ggplot(aes(cty,hwy)) +
    geom_count(show.legend = F, color = rgb(190,87,65, maxColorValue = 255)) +
    labs(
        title = "Counts plot",
        subtitle = "mpg: city vs highway mileage"
    ) +
    theme_bw()

ggsave("../output/plot3.pdf", width = 7, height = 4, dpi = 300)
```

# Divergent plot


```{r}
data4 <- read_csv("../data/data_plot4.csv")
glimpse(data4)
data4 |>
    mutate(
        mpg_z = scale(mpg),
        mpg_avg = if_else(mpg_z < 0, "Below Average", "Above Average")
    ) |> 
    ggplot(aes(x = mpg_z, y = reorder(car_name, mpg_z), fill = mpg_avg)) +
    geom_col() +
    labs(
        x = "MPG (Z-score)",
        y = "Car Name",
        fill = "Mileage"
    ) +
    scale_fill_manual(values = c(
        "Above Average" = rgb(164,190,160, maxColorValue = 255),
        "Below Average" = rgb(223,124,120, maxColorValue = 255)
    )) +
    theme_bw()

ggsave("../output/plot4.pdf", height = 6, width = 8)
```

# Lollipop chart


```{r}
data5 <- read_csv("../data/data_plot5.csv")
glimpse(data5)
data5 |> 
    ggplot(aes(x = reorder(manufacturer, mileage), y = mileage)) +
    geom_segment(aes(xend = reorder(manufacturer, mileage), y = 0, yend = mileage), color = "gray80") + 
    geom_point(color = "orange", size = 3) +
    scale_y_continuous(limits = c(0,25), breaks = seq(0,25,5)) +
    labs(
        title = "Lollipop Chart",
        subtitle = "Avgerage city mileage by manufacturer",
        y = "Avg. Mileage",
        x = "",
        caption = "source: mpg"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 75, hjust = 0.5, vjust = 0.5),
    panel.grid.major.x = element_blank())

ggsave("../output/plot5.pdf", height = 6, width = 8)
```

# Volcano


```{r}
library(ggrepel)
data6 <- read_csv("../data/data_plot6.csv")
glimpse(data6)
data6 |> 
    mutate(
        diffexp = case_when(
            pvalue < 0.05 & log2FoldChange < -0.6 ~ "down",
            pvalue < 0.05 & log2FoldChange > 0.6 ~ "up",
            .default = "no"
        ),
        label = if_else(diffexp %in% c("down", "up"), gene_symbol, NA)
    ) |> 
    ggplot(aes(x = log2FoldChange, y = -log10(pvalue), label = label, color = diffexp)) +
    geom_point(show.legend = F) +
    geom_text_repel(show.legend = F, color = "black") +
    scale_color_manual(values = c("down" = "blue", "no" = "black", "up" = "red")) +
    labs(
        title = "Volcano plot",
        y = "-log10 (P-value)",
        x = expression(log[2]*FC)
    ) +
    theme_classic() +
    theme(panel.background = element_rect(fill = "gray95", color = "gray80"))

ggsave("../output/plot6.pdf", height = 6, width = 8)
```